---
title: "Exploratory Data Analysis"
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\}, breaklines, breaknonspaceingroup, breakanywhere
        }
---

```{r}
library(haven)
library(tidyverse)
dat25 <- read_xpt("../data/LLCP2025.XPT")
dat24 <- read_xpt("../data/LLCP2024.XPT")
#dat23 <- read_xpt("../data/LLCP2023.XPT")
dat22 <- read_xpt("../data/LLCP2022.XPT")
#dat21 <- read_xpt("../data/LLCP2021.XPT")
#dat20 <- read_xpt("../data/LLCP2020.XPT")
dat19 <- read_xpt("../data/LLCP2019.XPT")
#dat18 <- read_xpt("../data/LLCP2018.XPT")
dat17 <- read_xpt("../data/LLCP2017.XPT")
```

# Exploratory data analysis

## variables asked only in 2025, 2024, 2022, 2019, and 2017

`NOBCUSE9` = What was main reason for not doing anything to keep you from getting pregnant? `PFPPRVN4` = The last time you had sexual intercourse, did you or your partner do anything to keep you from getting pregnant? () `FAMILY PLANNING` = The last time you had sexual intercourse, what did you or your partner do to keep you from getting pregnant? (lots of response categories!!!)

## variables asked in all surveys

`SEXVAR` = 2 if female

`MARITAL`= marital status (1=married, 2=divorced, 3=widowed, 4=separated, 5=never married,6=A member of an unmarried couple, 9 =refused)

`HOME`= homeowner status (1=own, 2=rent, 3=other arrangement, 7=don't know, 9=refused)

`PREGNANT` = current pregnancy status (1=yes, 2=no, 7=don't know, 9=refused)

`MENTHLTH` = Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good? (1-30, 88=none,77=don't know, 99=refused)

`EMPLOY` = employment status (1 = employed for wages, 2 = self-employed, 3 = out of work for 1 year or more, 4 = out of work for less than 1 year, 5 = homemaker, 6 = student, 7 = retired, 8 = unable to work, 9 = refused)

`INCOME` = income categories (1=less than 15k, 2=15k to 25k, 3= 25k to 35k, 4= 35k to 50k, 5=50k to 100k, 6=100k to 200k, 7=200k or more, 9=don't know)

`EDUCA` = What is the highest grade or year of school you completed? (1=no high school, 2=high school, 3=some college,4=college,9=don't know)

`INSURANCE` = Respondents aged 18-64 who have any form of health insurance (1=yes, 2=no, 9=don't know)

`RACE` = Race/ethnicity categories (1=nh white, 2=nh black, 3=nh native american, 4=nh asian, 5=nh native hawaiian, 7=nh multiracial,9=don't know)

## subset each survey year to include relevant variables, subsetting the sample to only women

Recode the states in each year since states that didn't collect enough data for a given year weren't included in the dataset

```{r}
sub_dat25 <- dat25 |> select(c("_STATE", "SEXVAR", "PREGNANT", "MENTHLTH", "EMPLOY1","_INCOMG1", "EDUCA", "_HCVU654", "_RACE", "RENTHOM1", "MARITAL", "TYPCNT10")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG1', INSURANCE='_HCVU654', EMPLOY='EMPLOY1', FAMILY_PLANNING="TYPCNT10", "HOME"=RENTHOM1, RACE="_RACE")

state_labels_2025 <- c(
  `1`  = "Alabama",
  `2`  = "Alaska",
  `4`  = "Arizona",
  `5`  = "Arkansas",
  `8`  = "Colorado",
  `9`  = "Connecticut",
  `10` = "Delaware",
  `11` = "District of Columbia",
  `12` = "Florida",
  `13` = "Georgia",
  `15` = "Hawaii",
  `16` = "Idaho",
  `17` = "Illinois",
  `18` = "Indiana",
  `19` = "Iowa",
  `20` = "Kansas",
  `21` = "Kentucky",
  `22` = "Louisiana",
  `23` = "Maine",
  `24` = "Maryland",
  `25` = "Massachusetts",
  `26` = "Michigan",
  `27` = "Minnesota",
  `29` = "Missouri",
  `30` = "Montana",
  `31` = "Nebraska",
  `33` = "New Hampshire",
  `34` = "New Jersey",
  `35` = "New Mexico",
  `36` = "New York",
  `37` = "North Carolina",
  `38` = "North Dakota",
  `39` = "Ohio",
  `40` = "Oklahoma",
  `41` = "Oregon",
  `42` = "Pennsylvania",
  `44` = "Rhode Island",
  `45` = "South Carolina",
  `46` = "South Dakota",
  `47` = "Tennessee",
  `48` = "Texas",
  `49` = "Utah",
  `50` = "Vermont",
  `51` = "Virginia",
  `53` = "Washington",
  `54` = "West Virginia",
  `55` = "Wisconsin",
  `56` = "Wyoming",
  `66` = "Guam",
  `72` = "Puerto Rico"
)

sub_dat25 <- sub_dat25 |>
  dplyr::mutate(
    `_STATE` = as.character(unname(
      state_labels_2025[as.character(`_STATE`)]
    )
  )) |>
    mutate(EDUCA = case_when(
        EDUCA %in% 1:3 ~ 1, # no high school
        EDUCA == 4 ~ 2, # high school
        EDUCA == 5 ~ 3 ,#Some college
        EDUCA == 6 ~ 4 ,#College graduate
        EDUCA == 9 ~ 9  #Don't know
  ))
```

```{r}
sub_dat24 <- dat24 |> select(c("_STATE","SEXVAR","PREGNANT", "MENTHLTH", "EMPLOY1","_INCOMG1", "EDUCA", "_HCVU654","_RACE", "RENTHOM1", "MARITAL", "TYPCNTR9")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG1', INSURANCE='_HCVU654', EMPLOY='EMPLOY1', FAMILY_PLANNING="TYPCNTR9", "HOME"=RENTHOM1, RACE="_RACE")

state_labels_2024 <- c(
  `1`  = "Alabama",
  `2`  = "Alaska",
  `4`  = "Arizona",
  `5`  = "Arkansas",
  `6`  = "California",
  `8`  = "Colorado",
  `9`  = "Connecticut",
  `10` = "Delaware",
  `11` = "District of Columbia",
  `12` = "Florida",
  `13` = "Georgia",
  `15` = "Hawaii",
  `16` = "Idaho",
  `17` = "Illinois",
  `18` = "Indiana",
  `19` = "Iowa",
  `20` = "Kansas",
  `21` = "Kentucky",
  `22` = "Louisiana",
  `23` = "Maine",
  `24` = "Maryland",
  `25` = "Massachusetts",
  `26` = "Michigan",
  `27` = "Minnesota",
  `28` = "Mississippi",
  `29` = "Missouri",
  `30` = "Montana",
  `31` = "Nebraska",
  `32` = "Nevada",
  `33` = "New Hampshire",
  `34` = "New Jersey",
  `35` = "New Mexico",
  `36` = "New York",
  `37` = "North Carolina",
  `38` = "North Dakota",
  `39` = "Ohio",
  `40` = "Oklahoma",
  `41` = "Oregon",
  `42` = "Pennsylvania",
  `44` = "Rhode Island",
  `45` = "South Carolina",
  `46` = "South Dakota",
  `48` = "Texas",
  `49` = "Utah",
  `50` = "Vermont",
  `51` = "Virginia",
  `53` = "Washington",
  `54` = "West Virginia",
  `55` = "Wisconsin",
  `56` = "Wyoming",
  `66` = "Guam",
  `72` = "Puerto Rico",
  `78` = "Virgin Islands"
)
sub_dat24 <- sub_dat24 |>
  dplyr::mutate(
    `_STATE` = as.character(unname(
      state_labels_2024[as.character(`_STATE`)]
    )
  )) |> mutate(EDUCA = case_when(
        EDUCA %in% 1:3 ~ 1, # no high school
        EDUCA == 4 ~ 2, # high school
        EDUCA == 5 ~ 3 ,#Some college
        EDUCA == 6 ~ 4 ,#College graduate
        EDUCA == 9 ~ 9  #Don't know
  ))
```

```{r}
# Roe v. Wade was overturned on June 24, 2022
sub_dat22 <- dat22 |> select(c("_STATE","SEXVAR","IMONTH","IDAY","IYEAR","PREGNANT", "MENTHLTH", "EMPLOY1","_INCOMG1", "EDUCA", "_HCVU652", "_RACE1", "TYPCNTR9", "RENTHOM1")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG1', INSURANCE='_HCVU652', EMPLOY='EMPLOY1', FAMILY_PLANNING="TYPCNTR9", HOME="RENTHOM1", RACE="_RACE1") 
state_labels_2022 <- c(
  `1`  = "Alabama",
  `2`  = "Alaska",
  `4`  = "Arizona",
  `5`  = "Arkansas",
  `6`  = "California",
  `8`  = "Colorado",
  `9`  = "Connecticut",
  `10` = "Delaware",
  `11` = "District of Columbia",
  `12` = "Florida",
  `13` = "Georgia",
  `15` = "Hawaii",
  `16` = "Idaho",
  `17` = "Illinois",
  `18` = "Indiana",
  `19` = "Iowa",
  `20` = "Kansas",
  `21` = "Kentucky",
  `22` = "Louisiana",
  `23` = "Maine",
  `24` = "Maryland",
  `25` = "Massachusetts",
  `26` = "Michigan",
  `27` = "Minnesota",
  `28` = "Mississippi",
  `29` = "Missouri",
  `30` = "Montana",
  `31` = "Nebraska",
  `32` = "Nevada",
  `33` = "New Hampshire",
  `34` = "New Jersey",
  `35` = "New Mexico",
  `36` = "New York",
  `37` = "North Carolina",
  `38` = "North Dakota",
  `39` = "Ohio",
  `40` = "Oklahoma",
  `41` = "Oregon",
  `42` = "Pennsylvania",
  `44` = "Rhode Island",
  `45` = "South Carolina",
  `46` = "South Dakota",
  `47` = "Tennessee",
  `48` = "Texas",
  `49` = "Utah",
  `50` = "Vermont",
  `51` = "Virginia",
  `53` = "Washington",
  `54` = "West Virginia",
  `55` = "Wisconsin",
  `56` = "Wyoming",
  `66` = "Guam",
  `72` = "Puerto Rico",
  `78` = "Virgin Islands"
)
sub_dat22 <- sub_dat22 |>
  dplyr::mutate(
    `_STATE` = as.character(unname(
      state_labels_2022[as.character(`_STATE`)]
    )
  ))|> mutate(EDUCA = case_when(
        EDUCA %in% 1:3 ~ 1, # no high school
        EDUCA == 4 ~ 2, # high school
        EDUCA == 5 ~ 3 ,#Some college
        EDUCA == 6 ~ 4 ,#College graduate
        EDUCA == 9 ~ 9  #Don't know
  ))

```

```{r}
sub_dat19 <- dat19 |> select(c("_STATE","SEXVAR","PREGNANT","MENTHLTH","EMPLOY1","_INCOMG", "EDUCA", "_HCVU651", "_RACE", "RENTHOM1", "MARITAL", "TYPCNTR8")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG', INSURANCE='_HCVU651', EMPLOY='EMPLOY1', FAMILY_PLANNING="TYPCNTR8", "HOME"=RENTHOM1, RACE="_RACE")
state_labels_2019 <- c(
  `1`  = "Alabama",
  `2`  = "Alaska",
  `4`  = "Arizona",
  `5`  = "Arkansas",
  `6`  = "California",
  `8`  = "Colorado",
  `9`  = "Connecticut",
  `10` = "Delaware",
  `11` = "District of Columbia",
  `12` = "Florida",
  `13` = "Georgia",
  `15` = "Hawaii",
  `16` = "Idaho",
  `17` = "Illinois",
  `18` = "Indiana",
  `19` = "Iowa",
  `20` = "Kansas",
  `21` = "Kentucky",
  `22` = "Louisiana",
  `23` = "Maine",
  `24` = "Maryland",
  `25` = "Massachusetts",
  `26` = "Michigan",
  `27` = "Minnesota",
  `28` = "Mississippi",
  `29` = "Missouri",
  `30` = "Montana",
  `31` = "Nebraska",
  `32` = "Nevada",
  `33` = "New Hampshire",
  `35` = "New Mexico",
  `36` = "New York",
  `37` = "North Carolina",
  `38` = "North Dakota",
  `39` = "Ohio",
  `40` = "Oklahoma",
  `41` = "Oregon",
  `42` = "Pennsylvania",
  `44` = "Rhode Island",
  `45` = "South Carolina",
  `46` = "South Dakota",
  `47` = "Tennessee",
  `48` = "Texas",
  `49` = "Utah",
  `50` = "Vermont",
  `51` = "Virginia",
  `53` = "Washington",
  `54` = "West Virginia",
  `55` = "Wisconsin",
  `56` = "Wyoming",
  `66` = "Guam",
  `72` = "Puerto Rico"
)
sub_dat19 <- sub_dat19 |>
  dplyr::mutate(
    `_STATE` = as.character(unname(
      state_labels_2019[as.character(`_STATE`)]
    )
  )) |>
    mutate(EDUCA = case_when(
        EDUCA %in% 1:3 ~ 1, # no high school
        EDUCA == 4 ~ 2, # high school
        EDUCA == 5 ~ 3 ,#Some college
        EDUCA == 6 ~ 4 ,#College graduate
        EDUCA == 9 ~ 9  #Don't know
  ))
```

```{r}
sub_dat17 <- dat17 |> select(c("_STATE", "SEX","PREGNANT", "MENTHLTH", "EMPLOY1", "_INCOMG", "EDUCA", "_HCVU651",  "_RACE", "RENTHOM1", "MARITAL", "TYPCNTR7")) |> filter(SEX==2) |> select(-SEX) |> rename(INCOME='_INCOMG', INSURANCE='_HCVU651', EMPLOY='EMPLOY1', FAMILY_PLANNING="TYPCNTR7", "HOME"=RENTHOM1, RACE="_RACE")
state_labels_2017 <- c(
  `1`  = "Alabama",
  `2`  = "Alaska",
  `4`  = "Arizona",
  `5`  = "Arkansas",
  `6`  = "California",
  `8`  = "Colorado",
  `9`  = "Connecticut",
  `10` = "Delaware",
  `11` = "District of Columbia",
  `12` = "Florida",
  `13` = "Georgia",
  `15` = "Hawaii",
  `16` = "Idaho",
  `17` = "Illinois",
  `18` = "Indiana",
  `19` = "Iowa",
  `20` = "Kansas",
  `21` = "Kentucky",
  `22` = "Louisiana",
  `23` = "Maine",
  `24` = "Maryland",
  `25` = "Massachusetts",
  `26` = "Michigan",
  `27` = "Minnesota",
  `28` = "Mississippi",
  `29` = "Missouri",
  `30` = "Montana",
  `31` = "Nebraska",
  `32` = "Nevada",
  `33` = "New Hampshire",
  `34` = "New Jersey",
  `35` = "New Mexico",
  `36` = "New York",
  `37` = "North Carolina",
  `38` = "North Dakota",
  `39` = "Ohio",
  `40` = "Oklahoma",
  `41` = "Oregon",
  `42` = "Pennsylvania",
  `44` = "Rhode Island",
  `45` = "South Carolina",
  `46` = "South Dakota",
  `47` = "Tennessee",
  `48` = "Texas",
  `49` = "Utah",
  `50` = "Vermont",
  `51` = "Virginia",
  `53` = "Washington",
  `54` = "West Virginia",
  `55` = "Wisconsin",
  `56` = "Wyoming",
  `66` = "Guam",
  `72` = "Puerto Rico"
)
sub_dat17 <- sub_dat17 |>
  dplyr::mutate(
    `_STATE` = as.character(unname(
      state_labels_2017[as.character(`_STATE`)]
    )
  )) |>
  mutate(EDUCA = case_when(
        EDUCA %in% 1:3 ~ 1, # no high school
        EDUCA == 4 ~ 2, # high school
        EDUCA == 5 ~ 3 ,#Some college
        EDUCA == 6 ~ 4 ,#College graduate
        EDUCA == 9 ~ 9  #Don't know
  ))

```

```{r,eval=FALSE}
sub_dat23 <- dat23 |> select(c("_STATE","SEXVAR","PREGNANT", "MENTHLTH","EMPLOY1", "_INCOMG1", "EDUCA", "_HCVU653", "_RACE")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG1', INSURANCE='_HCVU653', EMPLOY='EMPLOY1')


sub_dat21 <- dat21 |> select(c("_STATE","SEXVAR","PREGNANT", "MENTHLTH","EMPLOY1","_INCOMG1", "EDUCA","_HCVU652", "_RACE")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG1', INSURANCE='_HCVU652', EMPLOY='EMPLOY1') 

sub_dat20 <- dat20 |> select(c("_STATE","SEXVAR","PREGNANT", "MENTHLTH","EMPLOY1", "_INCOMG", "EDUCA",  "_HCVU651", "_RACE")) |> filter(SEXVAR==2) |> select(-SEXVAR) |> rename(INCOME='_INCOMG', INSURANCE='_HCVU651', EMPLOY='EMPLOY1')


sub_dat18 <- dat18 |> select(c("_STATE", "SEX1","PREGNANT", "MENTHLTH", "EMPLOY1", "_INCOMG", "EDUCA", "_HCVU651",  "_RACE", "RENTHOM1", "MARITAL")) |> filter(SEX1==2) |> select(-SEX1) |> rename(INCOME='_INCOMG', INSURANCE='_HCVU651', EMPLOY='EMPLOY1')
```

## merge relevant questionnaire years

```{r}
data_list <- list(
  `2017` = sub_dat17,
  `2019` = sub_dat19,
  `2022` = sub_dat22,
  `2024` = sub_dat24,
  `2025` = sub_dat25
)

vars_to_keep <- c(
  "_STATE",
  "PREGNANT",
  "MENTHLTH",
  "EMPLOY",
  "INCOME",
  "EDUCA",
  "INSURANCE",
  "RACE",
  "MARITAL",
  "HOME",
  "FAMILY_PLANNING",
  "IYEAR",
  "IMONTH",
  "IDAY"
)

analysis_data <- purrr::imap_dfr(
  data_list,
  function(dataset, year) {

    year <- as.integer(year)

    dataset <- dataset |>
      dplyr::select(dplyr::any_of(vars_to_keep)) |>
      dplyr::mutate(
        YEAR = year
      )

    if (year == 2022) {
      dataset <- dataset |>
        dplyr::mutate(
          IYEAR = as.integer(as.character(IYEAR)),
          IMONTH = as.integer(as.character(IMONTH)),
          IDAY = as.integer(as.character(IDAY)),

          date = as.Date(
            sprintf("%04d-%02d-%02d", IYEAR, IMONTH, IDAY)
          ),

          post_ban = dplyr::case_when(
            date < as.Date("2022-06-24") ~ 0L,
            date >= as.Date("2022-06-24") ~ 1L,
            TRUE ~ NA_integer_
          )
        )
    } else {
      dataset <- dataset |>
        dplyr::mutate(
          date = as.Date(NA),

          POST_BAN = dplyr::case_when(
            YEAR <= 2021 ~ 0L,
            YEAR >= 2023 ~ 1L,
            TRUE ~ NA_integer_
          )
        )
    }

    dataset
  }
) |>
  dplyr::rename(STATE = `_STATE`) |> select(-c(IYEAR, IMONTH, IDAY, date))


# check that post ban was correctly created
analysis_data |> filter(YEAR == 2025) |> janitor::tabyl(post_ban)
analysis_data |> filter(YEAR == 2024) |> janitor::tabyl(post_ban)
analysis_data |> filter(YEAR == 2022) |> janitor::tabyl(post_ban)
analysis_data |> filter(YEAR == 2019) |> janitor::tabyl(post_ban)
analysis_data |> filter(YEAR == 2017) |> janitor::tabyl(post_ban)
```

## create U.S. census regions

```{r}
library(tibble)

census_region_lookup <- tribble(
  ~STATE, ~REGION, ~DIVISION,

  # Northeast
  "Connecticut",          "Northeast", "New England",
  "Maine",                "Northeast", "New England",
  "Massachusetts",        "Northeast", "New England",
  "New Hampshire",        "Northeast", "New England",
  "Rhode Island",         "Northeast", "New England",
  "Vermont",              "Northeast", "New England",
  "New Jersey",           "Northeast", "Middle Atlantic",
  "New York",             "Northeast", "Middle Atlantic",
  "Pennsylvania",         "Northeast", "Middle Atlantic",

  # Midwest
  "Illinois",             "Midwest", "East North Central",
  "Indiana",              "Midwest", "East North Central",
  "Michigan",             "Midwest", "East North Central",
  "Ohio",                 "Midwest", "East North Central",
  "Wisconsin",            "Midwest", "East North Central",
  "Iowa",                 "Midwest", "West North Central",
  "Kansas",               "Midwest", "West North Central",
  "Minnesota",            "Midwest", "West North Central",
  "Missouri",             "Midwest", "West North Central",
  "Nebraska",             "Midwest", "West North Central",
  "North Dakota",         "Midwest", "West North Central",
  "South Dakota",         "Midwest", "West North Central",

  # South
  "Delaware",             "South", "South Atlantic",
  "District of Columbia", "South", "South Atlantic",
  "Florida",              "South", "South Atlantic",
  "Georgia",              "South", "South Atlantic",
  "Maryland",             "South", "South Atlantic",
  "North Carolina",       "South", "South Atlantic",
  "South Carolina",       "South", "South Atlantic",
  "Virginia",             "South", "South Atlantic",
  "West Virginia",        "South", "South Atlantic",
  "Alabama",              "South", "East South Central",
  "Kentucky",             "South", "East South Central",
  "Mississippi",          "South", "East South Central",
  "Tennessee",            "South", "East South Central",
  "Arkansas",             "South", "West South Central",
  "Louisiana",            "South", "West South Central",
  "Oklahoma",             "South", "West South Central",
  "Texas",                "South", "West South Central",

  # West
  "Arizona",              "West", "Mountain",
  "Colorado",             "West", "Mountain",
  "Idaho",                "West", "Mountain",
  "Montana",              "West", "Mountain",
  "Nevada",               "West", "Mountain",
  "New Mexico",           "West", "Mountain",
  "Utah",                 "West", "Mountain",
  "Wyoming",              "West", "Mountain",
  "Alaska",               "West", "Pacific",
  "California",           "West", "Pacific",
  "Hawaii",               "West", "Pacific",
  "Oregon",               "West", "Pacific",
  "Washington",           "West", "Pacific",

  # Territories: not assigned to a Census region/division
  "Guam",                 NA_character_, NA_character_,
  "Puerto Rico",          NA_character_, NA_character_,
  "Virgin Islands",       NA_character_, NA_character_
)

# remove Guam, Puerto Rico, and Virgin Islands from analytic sample
analysis_data_v2 <- analysis_data |>
  dplyr::left_join(
    census_region_lookup,
    by = "STATE"
  ) |> filter(!is.na(REGION)) 
```

```{r}
variable_labels <- list(

  MARITAL = c(
    `1` = "Married",
    `2` = "Divorced",
    `3` = "Widowed",
    `4` = "Separated",
    `5` = "Never married",
    `6` = "Member of an unmarried couple",
    `9` = "Refused"
  ),

  HOME = c(
    `1` = "Own",
    `2` = "Rent",
    `3` = "Other arrangement",
    `7` = "Don't know",
    `9` = "Refused"
  ),

  PREGNANT = c(
    `1` = "Yes",
    `2` = "No",
    `7` = "Don't know",
    `9` = "Refused"
  ),

  MENTHLTH = c(
    setNames(
      paste(1:30, ifelse(1:30 == 1, "day", "days")),
      as.character(1:30)
    ),
    `77` = "Don't know",
    `88` = "None",
    `99` = "Refused"
  ),

  EMPLOY = c(
    `1` = "Employed for wages",
    `2` = "Self-employed",
    `3` = "Out of work for 1 year or more",
    `4` = "Out of work for less than 1 year",
    `5` = "Homemaker",
    `6` = "Student",
    `7` = "Retired",
    `8` = "Unable to work",
    `9` = "Refused"
  ),

  INCOME = c(
    `1` = "Less than $15,000",
    `2` = "$15,000 to $24,999",
    `3` = "$25,000 to $34,999",
    `4` = "$35,000 to $49,999",
    `5` = "$50,000 to $99,999",
    `6` = "$100,000 to $199,999",
    `7` = "$200,000 or more",
    `9` = "Don't know"
  ),

  EDUCA = c(
    `1` = "No high school",
    `2` = "High school",
    `3` = "Some college",
    `4` = "College graduate",
    `9` = "Don't know"
  ),

  INSURANCE = c(
    `1` = "Yes",
    `2` = "No",
    `9` = "Don't know"
  ),

  RACE = c(
    `1` = "Non-Hispanic White",
    `2` = "Non-Hispanic Black",
    `3` = "Non-Hispanic Native American",
    `4` = "Non-Hispanic Asian",
    `5` = "Non-Hispanic Native Hawaiian",
    `6` = "Other race, non-Hispanic",
    `7` = "Non-Hispanic Multiracial",
    `8` = "Hispanic",
    `9` = "Don't know"
  )
)

variables_to_summarize <- c(
  "MARITAL",
  "HOME",
  "PREGNANT",
  "MENTHLTH",
  "EMPLOY",
  "INCOME",
  "EDUCA",
  "INSURANCE",
  "RACE"
)


```

## U.S. Population Characteristics Over Time

```{r}
# categorical variables
sapply(names(analysis_data_v2)[!names(analysis_data_v2) %in% c("STATE","YEAR","POST_BAN", "MENTHLTH")], function(var){
  ggplot(
  analysis_data_v2,
  aes(x = factor(.data[[var]]))
) +
  geom_bar(
    fill = "steelblue",
    color = "white"
  ) +
  facet_wrap(~ YEAR) +
  theme_minimal() +
  labs(
    x = var,
    y = "Count"
  )
})

# numerical variable
  ggplot(
  analysis_data_v2 ,
  aes(x = factor(MENTHLTH))
) +
  geom_bar(
    fill = "steelblue",
    color = "white"
  ) +
  facet_wrap(~ YEAR) +
  theme_minimal() +
  labs(
    title = "Distribution of Days with Poor Mental Health",
    x = var,
    y = "Count"
  )
```

```{r}
get_descriptive_stats_over_entire_period <- function(data, var) {
  
  tab <- data |>
    janitor::tabyl(
      {{ var }},
      show_na = TRUE
    )
  
  # The first column is the categorical variable
  category_col <- names(tab)[1]
  
  total_n <- sum(tab$n)
  
  valid_n <- sum(
    tab$n[!is.na(tab[[category_col]])]
  )
  
  tab |>
    dplyr::mutate(
      percent = 100 * n / total_n,
      valid_percent = dplyr::if_else(
        valid_n > 0 & !is.na(.data[[category_col]]),
        100 * n / valid_n,
        NA_real_
      )
    )
}
get_descriptive_stats_over_entire_period(analysis_data,PREGNANT)

get_descriptive_stats_over_entire_period(analysis_data,MENTHLTH)
get_descriptive_stats_over_entire_period(analysis_data,EMPLOY)
get_descriptive_stats_over_entire_period(analysis_data,INCOME)
get_descriptive_stats_over_entire_period(analysis_data,EDUCA)
get_descriptive_stats_over_entire_period(analysis_data,INSURANCE)
```

Around 66% of women report missing responses when asked if they are currently pregnant. 32% of women report they aren't currently pregnant.

## Trends in the U.S. Regions Over Time

```{r}
regional_summary <- analysis_data_v2 |>
  select(
    STATE,
    REGION,
    YEAR,
    all_of(variables_to_summarize)
  ) |>
  pivot_longer(
    cols = all_of(variables_to_summarize),
    names_to = "VARIABLE",
    values_to = "VALUE"
  ) |>
  mutate(
    VALUE_CODE = if_else(
      is.na(VALUE),
      "Missing",
      as.character(VALUE)
    ),

    VALUE_LABEL = map2_chr(
      VARIABLE,
      VALUE_CODE,
      function(variable, code) {
        if (code == "Missing") {
          return("Missing")
        }

        label <- unname(variable_labels[[variable]][code])

        if (is.na(label)) {
          paste("Unlabeled code:", code)
        } else {
          label
        }
      }
    )
  ) |>
  count(
    STATE,
    REGION,
    YEAR,
    VARIABLE,
    VALUE_CODE,
    VALUE_LABEL,
    name = "N"
  ) |>
  group_by(STATE, REGION, YEAR, VARIABLE) |>
  mutate(
    PERCENT = round(100 * N / sum(N),2)
  ) |>
  ungroup()

regional_summary


```

```{r}
plot_variable <- function(data, variable_name, plot_title, legend_title) {
  
  data |>
    dplyr::filter(VARIABLE == variable_name) |>
    ggplot(
      aes(
        x = factor(YEAR),
        y = PERCENT,
        fill = VALUE_LABEL
      )
    ) +
    geom_col(position = "dodge") +
    facet_wrap(~ REGION) +
    labs(
      title = plot_title,
      x = "Year",
      y = "Percent",
      fill = legend_title
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

var_by_region_plots <- list(
  marital = plot_variable(
    regional_summary,
    "MARITAL",
    "Marital Status by Region and Year",
    "Marital status"
  ),

  home = plot_variable(
    regional_summary,
    "HOME",
    "Homeowner Status by Region and Year",
    "Homeowner status"
  ),

  pregnant = plot_variable(
    regional_summary,
    "PREGNANT",
    "Pregnancy Status by Region and Year",
    "Pregnancy status"
  ),

  employ = plot_variable(
    regional_summary,
    "EMPLOY",
    "Employment Status by Region and Year",
    "Employment status"
  ),

  income = plot_variable(
    regional_summary,
    "INCOME",
    "Income Categories by Region and Year",
    "Income category"
  ),

  insurance = plot_variable(
    regional_summary,
    "INSURANCE",
    "Health Insurance Status by Region and Year",
    "Insurance status"
  ),

  race = plot_variable(
    regional_summary,
    "RACE",
    "Race and Ethnicity by Region and Year",
    "Race and ethnicity"
  )
)
```

```{r}
var_by_region_plots$marital
var_by_region_plots$home
var_by_region_plots$pregnant
var_by_region_plots$employ
var_by_region_plots$income
var_by_region_plots$insurance
var_by_region_plots$race
```
